---
title: "Analysis of tax-calculator output"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook: 
    df_print: paged
    fig_height: 6
    fig_width: 8
    toc: yes
    number_sections: yes
editor_options:
  chunk_output_type: inline
---


<!-- 

<style type="text/css">
div.main-container {
  max-width: 100%;
  margin-left: auto;
  margin-right: auto;
}
</style>

  html_document:
    fig_height: 6
    fig_width: 8
    toc: yes
    toc_depth: 5
editor_options:
  chunk_output_type: console or inline    
  
  html_document:
    toc: yes
    df_print: paged
  word_document:
    toc: yes
  pdf_document:
    toc: yes
    
-->


```{r setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(width = 150)
```



```{r libraries, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# source(here::here("includes", "libraries.r"))
library(tidyverse)
# if more than 60 rows, print 60 - enough for states
options(tibble.print_max = 60, tibble.print_min = 60) 
library(scales)
library(readxl)

library(arrow)

library(RColorBrewer)

library(maps)
library(usmap)
library(grid)
library(gridExtra)

library(btools) # Matt, get from my github page if needed
# library(bdata) # Matt, get from my github page if needed
library(ggrepel)
library(gt)
library(glue)
library(DT)
library(knitr)
library(kableExtra)
library(btools)
library(bdata)

devtools::session_info()

```


```{r define_runtype}
# pick one of the following
# runtype <- "stack_saltfirst"
# runtype <- "stack_jct"

```



```{r reform_names}

# map policy varnames and table column names

label_map <- read_csv("policyvar, colname
law2017, 2017 law
allrates2018, Rate reductions
sd2018, Standard deduction increase
persx2018, Personal exemption elimination
qbid2018_limitfalse, Qualified business income deduction
salt2018, SALT deduction cap
amt2018, Alternative minimum tax reductions
law2018, 2018 law
")

order_jct <- c("law2017", "allrates2018", "sd2018", "persx2018", "qbid2018_limitfalse", "salt2018", "amt2018", "law2018")
order_saltfirst <- c("law2017", "salt2018", "sd2018", "persx2018", "qbid2018_limitfalse", "amt2018", "allrates2018", "law2018")

label_map <- label_map %>%
  mutate(jct=match(policyvar, all_of(order_jct)) - 1,
         saltfirst=match(policyvar, all_of(order_saltfirst)) - 1) %>%
  mutate(fullname_jct=paste0(jct, "_", policyvar),
         fullname_salt=paste0(saltfirst, "_", policyvar))

label_map %>%
  arrange(saltfirst)

# which analysis to use? 
# policy_vars <- policy_vars_jct
# policy_colnames <- policy_colnames_jct
# policy_colnames <- str_sub(policystack, 3, -1) # fine as long as prefix is not greater than 9


```


```{r locations}
# define states of interest

tcout <- r"(C:\programs_python\puf_analysis\ignore\taxcalc_output\)"

jctdir <- paste0(tcout, "stack_jct/")
saltdir <- paste0(tcout, "stack_saltfirst/")
solodir <- paste0(tcout, "unstacked/")

solov2018dir <- paste0(tcout, "solo_vs_2018law/")

wtdir <- r"(C:\programs_python\puf_analysis\ignore\puf_versions\)"

wtfname <- "allweights2018_geo2017_grown.csv"


```


```{r constants}
  
idvars <- c("stack", "stabbr", "stname", "agi_stub", "agi_label")

# define some state sorts
# "AR", "CA", "CT", "FL", "MA", "NJ", "NY", "PA", "TX", "other"
usny_sort <- c("US", "NY", "AR", "CA", "CT", "FL", "MA", "NJ",  "PA", "TX", "other")
nyus_sort <- c("NY", "US", "AR", "CA", "CT", "FL", "MA", "NJ",  "PA", "TX", "other")

source_note <- "\nAuthors' analysis of a sample of tax returns constructed to be representative of states at 2018 income levels."

```


```{r functions_utilities}
get_stname <- function(stabbr){
  stname <- case_when(stabbr == "US" ~ "United States",
                      stabbr == "other" ~ "Other states combined",
                      stabbr %in% state.abb ~ state.name[match(stabbr, state.abb)],
                      TRUE ~ "ERROR")
  return(stname)
}
# get_stname(usny_sort)

get_stabbr <- function(stname){
  stabbr <- case_when(stname == "United States" ~ "US",
                      stname == "Other states combined" ~ "other",
                      stname %in% state.name ~ state.abb[match(stname, state.name)],
                      TRUE ~ "ERROR")
  return(stabbr)
}
# get_stabbr(get_stname(usny_sort))


html_wrap <- function(s, len){
  # gt uses <br> to break columns in html, rather than \n, so 
  # wrap lines and then replace \n with <br> for column headings
  wrapped <- str_wrap(s, 15)
  return(str_replace_all(wrapped, "\\n", "<br>"))
}


```



```{r functions_data}

pol_change <- function(wide, policy1, policy2, vname="iitax"){
  # generic function to take two policies and create a file with changes and percent changes
  p1name <- policy1
  p2name <- policy2
  data <- wide %>%
    filter(variable==all_of(vname)) %>%
    rename(policy1=all_of(policy1), policy2=all_of(policy2)) %>%
    mutate(p1name=p1name, p2name=p2name, 
           diff=policy2 - policy1,
           pdiff=ifelse(policy1 > 0, diff / policy1 * 100, NA_real_)) %>% 
    select(variable, p1name, p2name, all_of(idvars), policy1, policy2, diff, pdiff)
  return(data)
}


```


```{r functions_tables}
pol_change_table <- function(data, tab_title, tab_subtitle, p1name, p2name){
  stub <- names(data)[1]
  stub_label <- ifelse(stub=="stname", "", "Adjusted gross income")
  dollar_scale <- ifelse(stub=="stname", 1e-9, 1e-6)
  dollar_decimal <- ifelse(stub=="stname", 1, 0)
  # dollar_scale_label <- ifelse(stub=="stname", "millions", "billions")
  
  p1label <- html_wrap(p1name, 15)
  p2label <- html_wrap(p2name, 15)
  
  tab <- data %>%
    rename(stub=all_of(stub)) %>%
    gt() %>%  tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      stub = stub_label,
      policy1 = html(p1label),
      policy2 = html(p2label),
      diff = html("$ change"),
      pdiff = html("% change")
    ) %>%
    fmt_currency(
      columns = c("policy1", "policy2", "diff"),
      rows=1,
      decimals = dollar_decimal,
      scale_by = dollar_scale,
      suffixing = FALSE
    ) %>%
    fmt_number(
      columns = c("policy1", "policy2", "diff"),
      rows=2:nrow(data),
      decimals = dollar_decimal,
      scale_by = dollar_scale,
      suffixing = FALSE
    ) %>%
    fmt_percent(
      columns="pdiff",
      scale_values = FALSE,
      decimals = 1
    ) %>% 
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_source_note(source_note)
  return(tab)
}


pdiff_wide_table <- function(data, tab_title, tab_subtitle){
  data %>%
    gt() %>%  tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      agi_label = "Adjusted gross income"
    ) %>%
    fmt_percent(
      columns=2:ncol(data),
      scale_values = FALSE,
      decimals = 1
    ) %>% 
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_source_note(source_note)
}

diff_wide_table <- function(data, tab_title, tab_subtitle){
  data %>%
    gt() %>%  tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      agi_label = "Adjusted gross income"
    ) %>%
    fmt_currency(
      columns = 2:ncol(data),
      rows=1,
      decimals = 0,
      suffixing = FALSE
    ) %>%
    fmt_number(
      columns = 2:ncol(data),
      rows=2:nrow(data),
      decimals = 0,
      suffixing = FALSE
    ) %>%    
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_source_note(source_note)
}


```


```{r functions_state_byagi_table}
get_state_policies <- function(st, policy1, policy2){
  data <- df_changes_jct %>%
    filter(stabbr==st) %>%
    select(all_of(idvars), policy1=all_of(policy1), policy2=all_of(policy2), diff, pdiff) %>%
    arrange(agi_stub) %>%
    select(agi_label, policy1, policy2, diff, pdiff)
  return(data)
}

get_state_tab <- function(st, policy1, policy2, p1name, p2name, dollars){
  data <- get_state_policies(st, policy1, policy2)
  tab_title <- paste0(get_stname(st), ": Total federal income tax liability in 2018: ", p1name, ", and ", p2name)
  tab_subtitle <- paste0("Amounts are in ", dollars, " of 2018 dollars")
  tab <- pol_change_table(data, tab_title, tab_subtitle, p1name, p2name)
  return(tab)
}

```


# Get data

## stubs
```{r stubs, rows.print=13}
# , "filer", "ht2_stub", "ht2range"

# create collapsed stubs we will use for reporting purposes, combining lowest income ranges

# get needed variables from the first file, which we will then collapse
basedf <- read_parquet(paste0(jctdir, "0_law2017.parquet"),
                       col_select = all_of(c("pid", "filer", "c00100", "ht2_stub", "ht2range")))

ht2stubs <- basedf %>%
  select(ht2_stub, ht2range) %>%
  distinct() %>%
  arrange(ht2_stub)

# create collapsed stubs that combine ranges 1, 2, and 3
agi_bounds <- c(-Inf, 25e3, 50e3, 75e3, 100e3, 200e3, 500e3, 1e6, Inf)
agistubs <- ht2stubs %>%
  mutate(agi_stub = ifelse(ht2_stub <= 3, 1, ht2_stub - 2),
         agi_label=ifelse(ht2_stub <= 3, "Under $25,000", ht2range)) %>%
  select(agi_stub, agi_label) %>%
  distinct() %>%
  mutate(agi_ge=agi_bounds[1:(length(agi_bounds) - 1)],
         agi_lt=agi_bounds[2:length(agi_bounds)]) %>%
  add_row(agi_stub = 99,
          agi_label = "  Totals")
agistubs # these are the ranges we will use for the report

get_stub <- function(agi, agi_bounds){
  # get the income stub for a given agi
  cut(agi, agi_bounds, labels=FALSE, right=FALSE)
}

# get_stub(c(-1000, 0, 24999, 25e3, 1e9), agi_bounds) # check

# define agi stub for every person in 2017 base year
pid_stubs <- basedf %>%
  select(pid, filer, c00100) %>%
  mutate(agi_stub=get_stub(c00100, agi_bounds)) %>%
  arrange(pid)

```

## weights
```{r weights}
weights <- read_csv(paste0(wtdir, wtfname))

weights_long <- weights %>%
  rename("US"="weight") %>%
  select(-c(ht2_stub, geoweight_sum)) %>%
  pivot_longer(-pid, names_to = "stabbr", values_to = "weight")

# check that we have  weights for states we want to look at
sts <- unique(weights_long$stabbr)
setdiff_all(sts, nyus_sort)  # should be no difference

```


## stacked data
```{r reforms, rows.print=25}
# get ALL of the stacked reforms -- jct stacking and salt-first stacking
# for now keep a minimal list of information about each reform

get_file <- function(policy, policydir, keepvars){
  fname <- paste0(policydir, policy, ".parquet")
  stack <- case_when(str_detect(policydir, "stack_jct") ~ "jct",
                     str_detect(policydir, "stack_saltfirst") ~ "saltfirst",
                     TRUE ~ "ERROR")
  df <- read_parquet(fname, 
                     col_select = all_of(keepvars)) %>%
    mutate(policy=policy,
           stack=all_of(stack))
  df
}


keepvars <- c("pid", "c00100", "iitax")

# get the stack_jct policies
# label_map
# get_file(label_map$fullname_jct[1], jctdir, keepvars)
# get_file(label_map$fullname_salt[1], saltdir, keepvars)

policy_jct <- map_dfr(label_map$fullname_jct, get_file, jctdir, keepvars)
policy_saltfirst <- map_dfr(label_map$fullname_salt, get_file, saltdir, keepvars)
policydf <- bind_rows(policy_jct, policy_saltfirst)
count(policydf, stack, policy)

```

## unstacked data and produce a quick table of differences
```{r}
pnames <- list.files(solodir) %>% str_remove(".parquet")

get_solo <- function(policy, policydir, keepvars){
  fname <- paste0(policydir, policy, ".parquet")
  df <- read_parquet(fname, 
                     col_select = all_of(keepvars)) %>%
    mutate(policy=policy)
  df
}

# get_solo(pnames[1], solodir, c("pid", "s006", "c00100", "iitax"))
dfsolo <- map_dfr(pnames, get_solo, solodir, c("pid", "s006", "c00100", "iitax"))

# quick table of national differences
solosums <- dfsolo %>%
  group_by(policy) %>%
  summarise(iitax=sum(iitax * s006, na.rm=TRUE), .groups="drop")

solosums %>%
  filter(policy != "law2018_vs_law2017") %>%
  mutate(iitax=iitax / 1e9) %>%
  mutate(diff=iitax - iitax[policy=="law2017_vs_law2017"],
         pdiff = diff / iitax[policy=="law2017_vs_law2017"] * 100) %>%
  arrange(diff) %>%
  kable(digits=2, format.args=list(big.mark=","))

synergy <- solosums %>%
  filter(policy != "law2018_vs_law2017") %>%
  mutate(iitax=iitax / 1e9) %>%
  mutate(diff=iitax - iitax[policy=="law2017_vs_law2017"],
         pdiff = diff / iitax[policy=="law2017_vs_law2017"] * 100) %>%
  summarise(law2017=iitax[policy=="law2017_vs_law2017"],
            law2018xqlimit=iitax[policy=="law2018xQlimit_vs_law2017"],
            sumdiffs=sum(diff[!policy %in% c("law2017_vs_law2017", "law2018xQlimit_vs_law2017")])) %>%
  mutate(netdiff=law2018xqlimit - law2017,
         synergy=netdiff - sumdiffs)
synergy


```


## collapse unstacked data
```{r}
# make solo_wide table dfsolo

collapse_unstack <- weights_long %>%
  left_join(dfsolo %>% select(pid, c00100, iitax, policy),
            by="pid") %>%
  left_join(pid_stubs %>% select(-c00100), by = "pid") %>%
  group_by(policy, stabbr, agi_stub, filer) %>%
  summarise(c00100=sum(weight * c00100, na.rm=TRUE),
            iitax=sum(weight * iitax, na.rm=TRUE),
            .groups="drop")

agisums_unstack <- collapse_unstack %>%
  group_by(policy, stabbr, filer) %>%  
  summarise(c00100=sum(c00100, na.rm=TRUE),
            iitax=sum(iitax, na.rm=TRUE),
            .groups="drop") %>%
  mutate(agi_stub=99)

collapse_all_unstack <- bind_rows(collapse_unstack, agisums_unstack) %>%
  left_join(agistubs %>% select(agi_stub, agi_label), by="agi_stub") %>%
  mutate(stname=get_stname(stabbr)) %>%
  arrange(policy, stabbr, agi_stub)

# check filer and nonfiler sums
# collapse_all %>%
#   filter(stack=="jct", stabbr=="US", agi_stub==99) %>%
#   select(policy, filer, c00100, iitax) %>%
#   pivot_longer(cols=c(c00100, iitax)) %>%
#   mutate(filer=ifelse(filer, "filer", "nonfiler")) %>%
#   pivot_wider(names_from = "filer") %>%
#   mutate(nonfile_pct=nonfiler / (nonfiler + filer) * 100) %>%
#   arrange(name, -nonfile_pct) %>%
#   kable(digits=2, format.args=list(big.mark=","), caption="Non filers percent")
  

# this is our main analysis file
iitax_filers_unstack <- collapse_all_unstack %>%
  filter(filer) %>%
  select(-filer, -c00100) %>%
  arrange(stabbr, agi_stub, policy)

solo_wide <- collapse_all_unstack %>%
  filter(filer) %>%
  select(-filer, -c00100) %>%
  pivot_wider(names_from = policy, values_from = iitax)


```



## collapse stacked data and keep weights, too
```{r ONETIME_collapse}
# collapse by stack, policy, state, agi range
# this will be the main file that forms the basis of almost all analysis

collapse <- weights_long %>%
  left_join(policydf, by = "pid") %>%
  left_join(pid_stubs %>% select(-c00100), by = "pid") %>%
  group_by(stack, policy, stabbr, agi_stub, filer) %>%
  summarise(c00100=sum(weight * c00100, na.rm=TRUE),
            iitax=sum(weight * iitax, na.rm=TRUE),
            weight=sum(weight, na.rm=TRUE),
            .groups="drop")
summary(collapse)

saveRDS(collapse, here::here("data", "collapse.rds"))

```


```{r create_analysis_file, rows.print=20}
collapse <- readRDS(here::here("data", "collapse.rds"))

agisums <- collapse %>%
  group_by(stack, policy, stabbr, filer) %>%  
  summarise(c00100=sum(c00100, na.rm=TRUE),
            iitax=sum(iitax, na.rm=TRUE),
            weight=sum(weight, na.rm=TRUE),
            .groups="drop") %>%
  mutate(agi_stub=99)

collapse_all <- bind_rows(collapse, agisums) %>%
  left_join(agistubs %>% select(agi_stub, agi_label), by="agi_stub") %>%
  mutate(stname=get_stname(stabbr)) %>%
  arrange(stack, policy, stabbr, agi_stub)

# check filer and nonfiler sums
# collapse_all %>%
#   filter(stack=="jct", stabbr=="US", agi_stub==99) %>%
#   select(policy, filer, c00100, iitax) %>%
#   pivot_longer(cols=c(c00100, iitax)) %>%
#   mutate(filer=ifelse(filer, "filer", "nonfiler")) %>%
#   pivot_wider(names_from = "filer") %>%
#   mutate(nonfile_pct=nonfiler / (nonfiler + filer) * 100) %>%
#   arrange(name, -nonfile_pct) %>%
#   kable(digits=2, format.args=list(big.mark=","), caption="Non filers percent")
  

# this is our main analysis file
iitax_filers <- collapse_all %>%
  filter(filer) %>%
  select(-filer, -c00100) %>%
  arrange(stack, stabbr, agi_stub, policy) %>%
  group_by(stack, stabbr, stname, agi_stub) %>%
  mutate(change=iitax - lag(iitax))


# functions to make a wide tax file and a wide change file, for a given stacking order
iitax_filers
stack_keep <- "jct"

get_wide <- function(stack_keep){
  iitax_filers %>%
    filter(stack==stack_keep) %>%
    select(-change) %>%
    pivot_wider(names_from=policy, values_from=iitax) %>%
    arrange(stack, stabbr, agi_stub, agi_label)
}


# make a file of changes for each stacking order that walks from 2017 to 2018 law
get_changedf <- function(stack_keep){
  
  policies <- case_when(stack_keep == "jct" ~ label_map$fullname_jct,
                        stack_keep == "saltfirst" ~ label_map$fullname_salt,
                        TRUE ~ "ERROR") %>% sort()
  
  policies_use <- setdiff(policies, policies[1])
    
  changes <- iitax_filers %>% 
    filter(stack==stack_keep, policy %in% policies_use) %>%
    select(-iitax) %>%
    pivot_wider(names_from = policy, values_from = change) %>%
    rename(other=policies[length(policies)])
  
  change_order <- setdiff(names(changes), idvars)
  
  # now bring in the ends and compute total change
  ends <- iitax_filers %>% 
    filter(stack==stack_keep, policy %in% policies[c(1, length(policies))]) %>%
    select(-change) %>%
    pivot_wider(names_from = policy, values_from = iitax)
  
  start <- policies[1]
  end <- policies[length(policies)]
  
  changedf <- left_join(changes, ends, by=idvars) %>%
    select(all_of(idvars), policies[1], all_of(change_order), policies[length(policies)]) %>%
    mutate(diff=!!as.name(end) - !!as.name(start),
           pdiff=ifelse(!!as.name(start) > 0, diff / !!as.name(start) * 100, NA_real_)) %>%
    ungroup()
  return(changedf)
}


get_wide("jct")
get_wide("saltfirst")

get_changedf("jct")
get_changedf("saltfirst")

# create separate data frames of changes because they will not have the same columns
df_changes_jct <- get_changedf("jct")
df_changes_saltfirst <- get_changedf("saltfirst")


```

# Begin analysis
## Change from 2017 law to 2018 law

```{r}
# prep
mapdf <- label_map %>%
  arrange(jct) %>%
  select(jct, policyvar, colname, fullname=fullname_jct)
vars <- mapdf$fullname[c(1, nrow(mapdf))]

policy1 <- vars[1]
p1name <- "2017 law"

policy2 <- vars[2]
p2name <- "2018 law"


```


```{r 2018vs2017_summary, rows.print=20}

data <- df_changes_jct %>%
  filter(agi_stub==99) %>%
  select(idvars, policy1=all_of(policy1), policy2=all_of(policy2), diff, pdiff) %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr) %>%
  select(stname, policy1, policy2, diff, pdiff)

tab_title <- "Total federal income tax liability in 2018: 2017 law, and 2018 law"
tab_subtitle <- "Amounts are in billions of 2018 dollars"


tab <- pol_change_table(data, tab_title, tab_subtitle, p1name, p2name)
gtsave(tab, "2018vs2017_summary_table.png", path = here::here("results"))

tab

```


```{r taxcut_plot}
# data %>%
#   mutate(stabbr=get_stabbr(stname),
#          baseshare=policy1 / policy1[stname=="United States"] * 100,
#          cutshare=diff / diff[stname=="United States"] * 100) %>%
#   filter(!stabbr %in% c("US", "other")) %>%
#   ggplot(aes(baseshare, cutshare, label=stname))  +
#   geom_point() + 
#   geom_text()


# data %>%
#   mutate(stabbr=get_stabbr(stname),
#          baseshare=policy1 / policy1[stname=="United States"] * 100,
#          cutshare=diff / diff[stname=="United States"] * 100) %>%
#   filter(!stabbr %in% c("US", "other")) %>%
#   pivot_longer(cols=c(baseshare, cutshare)) %>%
#   ggplot(aes(x=stname, y=value, fill=name)) +
#     geom_bar(stat="identity", position=position_dodge())

p <- data %>%
  mutate(stabbr=get_stabbr(stname), pdiff=-pdiff,
         stname=ifelse(stabbr=="other", "States not\nshown, in\n aggregate", stname),
         stname=ifelse(stabbr=="US", "United\nStates", stname),
         grp=case_when(stabbr=="US" ~ 1,
                       stabbr=="NY" ~2,
                       TRUE ~ 3),
         grp=as.factor(grp)) %>%
  ggplot(aes(x=reorder(stname, -pdiff), y=pdiff, fill=grp)) +
    geom_bar(stat="identity") +
    scale_fill_manual(values=c("blue", "darkgreen", "lightgrey")) +
    scale_x_discrete(name=NULL) +
    scale_y_continuous(name="Tax cut (%)", breaks=seq(0, 20, 2.5)) +
    geom_hline(yintercept = -data$pdiff[data$stname=="United States"], size=0.1, colour="#525252") +
    theme_bw() +
    theme(legend.position = "none") +
    ggtitle("Tax cut in 2018 as percent of tax liability under 2017 law")
p
ggsave(filename=here::here("results", "taxcut_barchart.png"), plot=p, width=9, height=6, scale=1)

```



```{r 2018vs2017_one_state_byagi}

dollars <- "millions"

for(st in nyus_sort){
  print(st)
  tab <- get_state_tab(st, policy1, policy2, p1name, p2name, dollars)
  gtsave(tab, paste0("2018vs2017_byagi_table_", st, ".png"), path = here::here("results"))
  # print(tab)
}


```


```{r 2018vs2017_pdiff_byagi_wide, rows.print=20}

tab_title <- "Total federal income tax liability in 2018: 2017 law, and 2018 tax law"
tab_subtitle <- "Percent change from 2017 law to 2018 law"

pol_change <- function(wide, policy1, policy2){
  # generic function to take two policies and create a file with changes and percent changes
  p1name <- policy1
  p2name <- policy2
  data <- wide %>%
    filter(variable==all_of(vname)) %>%
    rename(policy1=all_of(policy1), policy2=all_of(policy2)) %>%
    mutate(p1name=p1name, p2name=p2name, 
           diff=policy2 - policy1,
           pdiff=ifelse(policy1 > 0, diff / policy1 * 100, NA_real_)) %>% 
    select(variable, p1name, p2name, all_of(idvars), policy1, policy2, diff, pdiff)
  return(data)
}
solo_wide

policy1 <- "law2017_vs_law2017"
p1name <- "2017 law"
policy2 <- "law2018xQlimit_vs_law2017"
p2name <- "2018 law"
data <- solo_wide %>%
  select(stabbr, stname, agi_stub, agi_label, policy1=all_of(policy1), policy2=all_of(policy2)) %>%
  mutate(p1name=p1name, p2name=p2name, 
           diff=policy2 - policy1,
           pdiff=ifelse(policy1 > 0, diff / policy1 * 100, NA_real_)) %>% 
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr, agi_stub) %>%
  select(agi_label, stname, pdiff) %>%
  pivot_wider(names_from = stname, values_from = pdiff)


tab <- pdiff_wide_table(data, tab_title, tab_subtitle)

gtsave(tab, "2018vs2017_pdiff_byagi_wide_table.png", path = here::here("results"))
tab

```


```{r pdiff_agirange_scatter}

pdata <- data %>%
  left_join(agistubs %>% select(agi_stub, agi_label), by="agi_label") %>%
  pivot_longer(-c(agi_stub, agi_label)) %>%
  filter(!agi_stub %in% c(1, 2, 99)) %>%
  mutate(stabbr=get_stabbr(name),
         agi_label=str_wrap(agi_label, 12),
         value=-value,
         grp=case_when(stabbr=="US" ~ 1,
                       stabbr=="NY" ~ 2,
                       TRUE ~ 3),
         grp=as.factor(grp)) %>%
  arrange(agi_stub, -value)

nudgex <- function(){
  df <- pdata %>%
    mutate(nudgex=case_when(agi_stub==3 & stabbr=="other" ~ -.1,
                            agi_stub==3 & stabbr=="CT" ~ -.075,
                            agi_stub==4 & stabbr %in% c("TX", "FL") ~ -0.075,
                            agi_stub==5 & stabbr=="other" ~ -.1,
                            agi_stub==5 & stabbr=="CT" ~ -.075,
                            agi_stub==6 & stabbr=="TX" ~ -.075,
                            agi_stub==6 & stabbr=="PA" ~ -.075,
                            agi_stub==6 & stabbr=="CA" ~ -.075,
                            agi_stub==6 & stabbr=="NJ" ~ -.075,
                            agi_stub==7 & stabbr=="FL" ~ -.075,
                            agi_stub==7 & stabbr=="PA" ~ -.075,
                            TRUE ~ .075))
  return(df$nudgex)
}

nudgey <- function(){
  df <- pdata %>%
    mutate(nudgey=case_when(agi_stub==3 & stabbr %in% c("US") ~ -0.25,
                            agi_stub==4 & stabbr %in% c("FL") ~ -0.25,
                            TRUE ~ 0.0))
  return(df$nudgey)
}
# nudgey()

p <- pdata %>%
  ggplot(aes(x=reorder(agi_label, agi_stub), y=value, label=stabbr, colour=grp)) +
    geom_point(size=0.5) +
    # geom_text(size=2, nudge_x=0.15, check_overlap = TRUE) +
    geom_text(fontface="bold", nudge_x=nudgex(), nudge_y=nudgey(), size=1.75, check_overlap = TRUE) +
    # geom_text_repel(size=2, nudge_x=0.15) +
    geom_hline(yintercept = 0, colour="red", size=0.1) +
    scale_colour_manual(values=c("blue", "darkgreen", "#525252")) +
    scale_x_discrete(name=NULL) +
    scale_y_continuous(name="Tax cut (%)") +
    ggtitle("Tax cut in 2018 as percent of liability under 2017 law",
            subtitle="By adjusted gross income range") +
    theme_bw() +
    theme(legend.position = "none") +
    labs(caption = "\nLowest income group had very large percentage reductions and is excluded to improve readability.") +
    theme(plot.caption = element_text(hjust = 0, size = 8, colour="#525252"))
ggsave(filename=here::here("results", "taxcut_agirange_scatter.png"), plot=p, width=9, height=6, scale=1)
# position=position_jitter(width=1,height=1)
p

```

## Change in average tax by agi range
```{r rows.print=20}
collapse_all 
count(collapse_all, stack, policy)

# , agi_stub==99, stabbr=="US"
avgtax <- collapse_all %>%
  filter(filer, stack=="jct", policy %in% c("0_law2017", "7_law2018")) %>%
  mutate(avgtax=iitax / weight,
         policy=case_when(policy=="0_law2017" ~ "policy1",
                          policy=="7_law2018" ~ "policy2",
                          TRUE ~ "ERROR")) %>%
  select(policy, stabbr, stname, agi_stub, agi_label, avgtax) %>%
  pivot_wider(names_from = policy, values_from = avgtax) %>%
  mutate(diff=policy2 - policy1,
         pdiff=ifelse(policy1 > 0, diff / policy1 * 100, NA_real_))
avgtax


```


```{r avgtax_change}

data <- avgtax %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr, agi_stub) %>%
  select(agi_label, stname, diff) %>%
  pivot_wider(names_from = stname, values_from = diff)

tab_title <- "Average federal income tax liability, per return, in 2018: 2017 law, and 2018 tax law"
tab_subtitle <- "Change in average tax, in dollars, from 2017 law to 2018 law"
tab <- diff_wide_table(data, tab_title, tab_subtitle)
gtsave(tab, "2018vs2017_dollardiff_byagi_wide_table.png", path = here::here("results"))
```


```{r avgtax_2017}
data <- avgtax %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr, agi_stub) %>%
  select(agi_label, stname, policy1) %>%
  pivot_wider(names_from = stname, values_from = policy1)

tab_title <- "Average federal income tax liability, per return, in 2018 under 2017 law"
tab <- diff_wide_table(data, tab_title, tab_subtitle=NULL)
gtsave(tab, "2017_dollarliab_byagi_wide_table.png", path = here::here("results"))
```


```{r avgtax_2018}
data <- avgtax %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr, agi_stub) %>%
  select(agi_label, stname, policy2) %>%
  pivot_wider(names_from = stname, values_from = policy2)

tab_title <- "Average federal income tax liability, per return, in 2018 under 2018 law"
tab <- diff_wide_table(data, tab_title, tab_subtitle=NULL)
gtsave(tab, "2018_dollarliab_byagi_wide_table.png", path = here::here("results"))
```



## Individual reforms measured in isolation

```{r rows.print=20}
solo_wide
ns(solo_wide)
label_map

# define the order of columns to match the JCT order
cols_order <- read_csv("policyvar, solovar
law2017, law2017_vs_law2017
allrates2018, allrates2018_vs_law2017
sd2018, sd2018_vs_law2017
persx2018, persx2018_vs_law2017
qbid2018_limitfalse, qbid2018_limitfalse_vs_law2017
salt2018, salt2018_vs_law2017
amt2018, amt2018_vs_law2017
other, other_2018vs2017_vs_law2017
law2018, law2018xQlimit_vs_law2017
")
cols_order <- cols_order %>%
   mutate(order=row_number()) %>%
  left_join(label_map %>% select(policyvar, colname),
            by="policyvar") %>%
  mutate(colname=ifelse(policyvar=="other", "Other provisions", colname)) %>%
  arrange(order)

diff_names <- cols_order %>%
  filter(!policyvar %in% c("law2017", "law2018")) %>%
  .$solovar

data <- solo_wide %>%
  filter(agi_stub==99) %>%
  select(stabbr, stname, agi_stub, agi_label, all_of(cols_order$solovar)) %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr, agi_stub) %>%
  mutate(across(.cols=all_of(diff_names),
                .fns= ~.x - law2017_vs_law2017))

tab_title <- "Changes from 2017 law for major provisions of the TCJA, measured in isolation"
tab_subtitle <- "Millions of dollars"

# create a list of labels for columns
tmp <- cols_order %>%
  filter(solovar %in% diff_names)
lablist <- split(tmp$colname, tmp$solovar)

# lablist["sd2018_vs_law2017"]

tab <- data %>%
  select(stname, all_of(diff_names)) %>%
    gt() %>%  tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      stname=""
    ) %>%
  cols_label(.list=lablist) %>%
    fmt_currency(
      columns = diff_names,
      rows=1,
      decimals = 0,
      scale_by = 1e-6,
      suffixing = FALSE
    ) %>%
    fmt_number(
      columns = diff_names,
      rows=2:nrow(data),
      decimals = 0,
      scale_by = 1e-6,
      suffixing = FALSE
    ) %>%
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_source_note(source_note)
  
gtsave(tab, "2018vs2017_provisions_bystate_isolation.png", path = here::here("results"))
tab

cols_order$colname

```



## Individual reforms estimated in the JCT order

## Changes by state
```{r}
glimpse(df_changes_jct)


tab_title <- "Changes from 2017 law for major provisions of the TCJA, measured in sequence"
tab_subtitle <- "Millions of dollars"

# create a list of labels for columns
tmp <- label_map %>%
  filter(fullname_jct %in% names(df_changes_jct)) %>%
  add_row(fullname_jct="other", colname="All other provisions combined")
lablist <- split(tmp$colname, tmp$fullname_jct)

data <- df_changes_jct %>%
  filter(agi_stub==99) %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr) %>%
  select(stname, all_of(label_map$fullname_jct), other, diff, pdiff) %>%
  relocate(other, .before = `7_law2018`)
  

tab <- data %>%
    gt() %>%  tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      stname="",
      diff="Total change",
      pdiff="% change"
    ) %>%
  cols_label(.list=lablist) %>%
    fmt_currency(
      columns = c(tmp$fullname_jct, "diff"),
      rows=1,
      decimals = 0,
      scale_by = 1e-6,
      suffixing = FALSE
    ) %>%
    fmt_number(
      columns = c(tmp$fullname_jct, "diff"),
      rows=2:nrow(data),
      decimals = 0,
      scale_by = 1e-6,
      suffixing = FALSE
    ) %>%
    fmt_percent(
      columns="pdiff",
      scale_values = FALSE,
      decimals = 1
    ) %>% 
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
  tab_spanner(
      label = "Tax change relative to 2018 law, estimated in left-to-right order, attributable to:",
      columns = c(setdiff(label_map$fullname_jct, c("0_law2017", "7_law2018")), "other")
    ) %>%
    tab_source_note(source_note)
  
gtsave(tab, "2018vs2017_provisions_bystate_jctstack.png", path = here::here("results"), vwidth=2000, vheight=2000)
tab
# vwidth = 992,   vheight = 744,

```


## Changes by income range

```{r us_changes}
glimpse(df_changes_jct)


tab_title <- "United States: Changes from 2017 law for major provisions of the TCJA, measured in sequence"
tab_subtitle <- "Impact by income range, millions of dollars"

# create a list of labels for columns
tmp <- label_map %>%
  filter(fullname_jct %in% names(df_changes_jct)) %>%
  add_row(fullname_jct="other", colname="All other provisions combined")
lablist <- split(tmp$colname, tmp$fullname_jct)

data <- df_changes_jct %>%
  filter(stabbr=="US") %>%
  arrange(agi_stub) %>%
  select(agi_label, all_of(label_map$fullname_jct), other, diff, pdiff) %>%
  relocate(other, .before = `7_law2018`)

tab <- data %>%
    gt() %>%  tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      agi_label="Adjusted gross income",
      diff="Total change",
      pdiff="% change"
    ) %>%
  cols_label(.list=lablist) %>%
    fmt_currency(
      columns = c(tmp$fullname_jct, "diff"),
      rows=1,
      decimals = 0,
      scale_by = 1e-6,
      suffixing = FALSE
    ) %>%
    fmt_number(
      columns = c(tmp$fullname_jct, "diff"),
      rows=2:nrow(data),
      decimals = 0,
      scale_by = 1e-6,
      suffixing = FALSE
    ) %>%
    fmt_percent(
      columns="pdiff",
      scale_values = FALSE,
      decimals = 1
    ) %>% 
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
  tab_spanner(
      label = "Tax change relative to 2018 law, estimated in left-to-right order, attributable to:",
      columns = c(setdiff(label_map$fullname_jct, c("0_law2017", "7_law2018")), "other")
    ) %>%
    tab_source_note(source_note)
  
gtsave(tab, "2018vs2017_provisions_us_byagi_jctstack.png", path = here::here("results"), vwidth=2000, vheight=2000)
tab
# vwidth = 992,   vheight = 744,

```


```{r ny_changes}
glimpse(df_changes_jct)


tab_title <- "New York: Changes from 2017 law for major provisions of the TCJA, measured in sequence"
tab_subtitle <- "Impact by income range, millions of dollars"

# create a list of labels for columns
tmp <- label_map %>%
  filter(fullname_jct %in% names(df_changes_jct)) %>%
  add_row(fullname_jct="other", colname="All other provisions combined")
lablist <- split(tmp$colname, tmp$fullname_jct)

data <- df_changes_jct %>%
  filter(stabbr=="NY") %>%
  arrange(agi_stub) %>%
  select(agi_label, all_of(label_map$fullname_jct), other, diff, pdiff) %>%
  relocate(other, .before = `7_law2018`)

tab <- data %>%
    gt() %>%  tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      agi_label="Adjusted gross income",
      diff="Total change",
      pdiff="% change"
    ) %>%
  cols_label(.list=lablist) %>%
    fmt_currency(
      columns = c(tmp$fullname_jct, "diff"),
      rows=1,
      decimals = 0,
      scale_by = 1e-6,
      suffixing = FALSE
    ) %>%
    fmt_number(
      columns = c(tmp$fullname_jct, "diff"),
      rows=2:nrow(data),
      decimals = 0,
      scale_by = 1e-6,
      suffixing = FALSE
    ) %>%
    fmt_percent(
      columns="pdiff",
      scale_values = FALSE,
      decimals = 1
    ) %>% 
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
  tab_spanner(
      label = "Tax change relative to 2018 law, estimated in left-to-right order, attributable to:",
      columns = c(setdiff(label_map$fullname_jct, c("0_law2017", "7_law2018")), "other")
    ) %>%
    tab_source_note(source_note)
  
gtsave(tab, "2018vs2017_provisions_nybyagi_jctstack.png", path = here::here("results"), vwidth=2000, vheight=2000)
tab
# vwidth = 992,   vheight = 744,

```


## Removing items from 2018 law, in isolation - how does it affect tax?

```{r}
# get the data
(files <- list.files(solov2018dir))

f <- function(file, dir, keepvars){
  policy <- str_remove(file, ".parquet")
  print(policy)
  path <- paste0(dir, file)
  df <- read_parquet(path, 
                     col_select = all_of(keepvars)) %>%
    mutate(policy=policy)
  df
}

keepvars <- c("pid", "c00100", "iitax")
policy_vs2018 <- map_dfr(files, f, solov2018dir, keepvars)

count(policy_vs2018, policy)
# [1] "amt2017_vs_law2018xQlimit"
# [1] "law2017_vs_law2018xQlimit"
# [1] "law2018xQlimit_vs_law2018xQlimit"
# [1] "persx2017_vs_law2018xQlimit"
# [1] "salt2017_vs_law2018xQlimit"
# [1] "sd2017_vs_law2018xQlimit"

```


```{r}
# prep the data
collapse_v2018 <- weights_long %>%
  left_join(policy_vs2018, by = "pid") %>%
  left_join(pid_stubs %>% select(-c00100), by = "pid") %>%
  group_by(policy, stabbr, agi_stub, filer) %>%
  summarise(c00100=sum(weight * c00100, na.rm=TRUE),
            iitax=sum(weight * iitax, na.rm=TRUE),
            weight=sum(weight, na.rm=TRUE),
            .groups="drop")
summary(collapse_v2018)

saveRDS(collapse_v2018, here::here("data", "collapse_v2018.rds"))

```


```{r}
# add sums, get pdiffs
collapse_v2018 <- readRDS(here::here("data", "collapse_v2018.rds"))

sums <- collapse_v2018 %>%
  group_by(policy, stabbr, filer) %>%  
  summarise(c00100=sum(c00100, na.rm=TRUE),
            iitax=sum(iitax, na.rm=TRUE),
            .groups="drop") %>%
  mutate(agi_stub=99)

sums_all <- bind_rows(collapse_v2018, sums) %>%
  left_join(agistubs %>% select(agi_stub, agi_label), by="agi_stub") %>%
  mutate(stname=get_stname(stabbr)) %>%
  arrange(policy, stabbr, agi_stub)

# verify that we see familiar numbers
sums_all %>% 
  filter(stabbr=="US", agi_stub==99, filer) %>%
  select(policy, iitax) %>%
  mutate(iitax=iitax / 1e9)
  

# create our main analysis file
# base_policy <- "law2018xQlimit_vs_law2018xQlimit"  # 2018 law with qbid ded limit false
base_policy <- "law2017_vs_law2018xQlimit"  # this is simply 2017 law
v2018_analysis <- sums_all %>%
  filter(filer) %>%
  select(-filer, -c00100) %>%
  arrange(stabbr, agi_stub, policy) %>%
  group_by(stabbr, agi_stub) %>%
  mutate(base_policy = iitax[policy==base_policy],
         diff=iitax - base_policy,
         pdiff=ifelse(base_policy > 0, diff / base_policy * 100, NA_real_)) %>%
  ungroup()
glimpse(v2018_analysis)

# create file with multiple policies vs 2017 law
# compare each policy to 2017 law

policy_2017 <- "law2017_vs_law2018xQlimit"
policy_2018 <- "law2018xQlimit_vs_law2018xQlimit"
v2018_analysis %>%
  filter(stabbr=="US", agi_stub==99) %>%
  select(policy, weight, iitax, base_policy, diff, pdiff)


```


```{r rows.print=30}
# create table that compares different policies for NY to US values
gaps <- v2018_analysis %>%
  filter(stabbr %in% c("US", "NY")) %>%
  select(-c(weight, base_policy, stname, iitax, diff)) %>%
  pivot_wider(names_from = stabbr, values_from = pdiff) %>%
  mutate(gap=NY - US) %>%
  arrange(agi_stub, policy)

gaps %>% filter(agi_stub==99)

# create data for table
# tcja, x salt, x sd, x amt (in isolation)
pols <- c("law2018xQlimit_vs_law2018xQlimit", "salt2017_vs_law2018xQlimit", "sd2017_vs_law2018xQlimit",
          "amt2017_vs_law2018xQlimit")
data <- gaps %>%
  filter(policy %in% pols) %>%
  pivot_longer(cols=c(NY, US, gap)) %>%
  # put vars in desired order
  mutate(name=factor(name, levels=c("US", "NY", "gap")),
         policy=factor(policy, levels=pols)) %>%
  arrange(policy, name) %>%
  unite("newcol", policy, name) %>%
  pivot_wider(names_from = newcol) %>%
  arrange(agi_stub) %>%
  select(-agi_stub)

col_names <- setdiff(names(data), "agi_label")
col_labels <- rep(c("US", "NY", "gap"), 4)
cbind(col_names, col_labels)

# create a list of labels for columns
lablist <- split(col_labels, col_names)

#  col_names                              col_labels
#  [1,] "law2018xQlimit_vs_law2018xQlimit_US"  "US"      
#  [2,] "law2018xQlimit_vs_law2018xQlimit_NY"  "NY"      
#  [3,] "law2018xQlimit_vs_law2018xQlimit_gap" "gap"     
#  [4,] "salt2017_vs_law2018xQlimit_US"        "US"      
#  [5,] "salt2017_vs_law2018xQlimit_NY"        "NY"      
#  [6,] "salt2017_vs_law2018xQlimit_gap"       "gap"     
#  [7,] "sd2017_vs_law2018xQlimit_US"          "US"      
#  [8,] "sd2017_vs_law2018xQlimit_NY"          "NY"      
#  [9,] "sd2017_vs_law2018xQlimit_gap"         "gap"     
# [10,] "amt2017_vs_law2018xQlimit_US"         "US"      
# [11,] "amt2017_vs_law2018xQlimit_NY"         "NY"      
# [12,] "amt2017_vs_law2018xQlimit_gap"        "gap"

tab_title <- "TCJA and TCJA with selected provisions removed"
tab_subtitle <- "Percent change in federal income tax liability from 2017 law, in 2018"
tab <- data %>%
    gt() %>%  tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      agi_label = "Adjusted gross income"
    ) %>%
    cols_label(.list=lablist) %>%
    fmt_percent(
      columns = 2:ncol(data),
      scale_values = FALSE,
      decimals = 1
    ) %>%    
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_spanner(
      label = html("Tax Cuts<br>and<br>Jobs Act"),
      columns = 2:4
    ) %>%
    tab_spanner(
      label = html("TCJA<br>without<br>SALT cap"),
      columns = 5:7
    ) %>%
    tab_spanner(
      label = html("TCJA<br>without standard<br>deduction increase"),
      columns = 8:10
    ) %>%
    tab_spanner(
      label = html("TCJA<br>without<br>AMT reductions"),
      columns = 11:13
    ) %>%
    tab_style(
      style = list(
        cell_borders(sides = "right", color = "#000000", style = "solid", weight = px(0.5))
      ),
      locations =list(
        cells_body(columns = c(4, 7, 10)),
        cells_column_labels(columns = c(4, 7, 10))
        )
      ) %>%
    cols_width(
      var(2:13) ~ pct(7)
    ) %>%
    tab_source_note(source_note)

gtsave(tab, "2018vs2017_selected_TCJAprovisions_removed.png", path = here::here("results"))
tab



```


```{r}
# just salt removed
pols <- c("law2018xQlimit_vs_law2018xQlimit", "salt2017_vs_law2018xQlimit")
data <- gaps %>%
  filter(policy %in% pols) %>%
  pivot_longer(cols=c(NY, US, gap)) %>%
  # put vars in desired order
  mutate(name=factor(name, levels=c("US", "NY", "gap")),
         policy=factor(policy, levels=pols)) %>%
  arrange(policy, name) %>%
  unite("newcol", policy, name) %>%
  pivot_wider(names_from = newcol) %>%
  arrange(agi_stub) %>%
  select(-agi_stub)

col_names <- setdiff(names(data), "agi_label")
col_labels <- rep(c("US", "NY", "gap"), 2)
cbind(col_names, col_labels)

# create a list of labels for columns
lablist <- split(col_labels, col_names)


tab_title <- "TCJA and TCJA with SALT cap removed"
tab_subtitle <- "Percent change in federal income tax liability from 2017 law, in 2018"
tab <- data %>%
    gt() %>%  tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      agi_label = "Adjusted gross income"
    ) %>%
    cols_label(.list=lablist) %>%
    fmt_percent(
      columns = 2:ncol(data),
      scale_values = FALSE,
      decimals = 1
    ) %>%    
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_spanner(
      label = html("Tax Cuts<br>and<br>Jobs Act"),
      columns = 2:4
    ) %>%
    tab_spanner(
      label = html("TCJA<br>without<br>SALT cap"),
      columns = 5:7
    ) %>%
    tab_style(
      style = list(
        cell_borders(sides = "right", color = "#000000", style = "solid", weight = px(0.5))
      ),
      locations =list(
        cells_body(columns = c(4)),
        cells_column_labels(columns = c(4))
        )
      ) %>%
    # cols_width(
    #   var(2:13) ~ pct(7)
    # ) %>%
    tab_source_note(source_note)
gtsave(tab, "2018vs2017_SALTcap_removed.png", path = here::here("results"))
tab

```



```{r}
# create table for removing salt from tcja
data <- v2018_analysis %>%
  filter(policy=="salt2017_vs_law2018xQlimit") %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr, agi_stub) %>%
  select(stname, agi_label, pdiff) %>%
  pivot_wider(names_from = stname, values_from = pdiff)


tab_title <- "abc"
tab_subtitle <- "def"
pdiff_wide_table(data, tab_title, tab_subtitle)


```





## Impact of uncapped SALT vs. 2018 law

```{r}
policy1 <- "law2018"
p1name <- "2018 law"

policy2 <- "law2018_xSALT"
p2name <- "2018 law with SALT uncapped"

```


```{r 2018vsSALT2017_summary, rows.print=20}

tab_title <- "Total federal income tax liability: 2018 law, and 2018 law with SALT uncapped"
tab_subtitle <- "Amounts are in billions of 2018 dollars"

data <- pol_change(allsums_wide, policy1, policy2) %>%
  filter(agi_stub==99) %>%
  select(all_of(idvars), policy1, policy2, diff, pdiff) %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr) %>%
  select(stname, policy1, policy2, diff, pdiff)

tab <- pol_change_table(data, tab_title, tab_subtitle, p1name, p2name)
gtsave(tab, "2018vsSALT2017_summary_table.png", path = here::here("results"))

tab

```


```{r 2018vsSALT2017_one_state_byagi}

dollars <- "millions"

for(st in c("US", "NY")){  # nyus_sort
  print(st)
  tab <- get_state_tab(st, policy1, policy2, p1name, p2name, dollars)
  gtsave(tab, paste0("2018vsSALT2017_byagi_table_", st, ".png"), path = here::here("results"))
  # print(tab)
}


```


```{r 2018vsSALT2017_pdiff_byagi_wide, rows.print=20}

tab_title <- "Total federal income tax liability: 2018 law, and 2018 law with SALT uncapped"
tab_subtitle <- "Percent change from 2018 law to 2018 law with SALT uncapped"

data <- pol_change(allsums_wide, policy1, policy2) %>%
  select(all_of(idvars), agi_stub, agi_label, stname, pdiff) %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr, agi_stub) %>%
  select(agi_label, stname, pdiff) %>%
  pivot_wider(names_from = stname, values_from = pdiff)

tab <- pdiff_wide_table(data, tab_title, tab_subtitle)

gtsave(tab, "2018vsSALT2017_pdiff_byagi_wide_table.png", path = here::here("results"))
tab


```


## Impact of the standard deduction increase, relative to TCJA

```{r}
policy1 <- "law2018"
p1name <- "2018 law"

policy2 <- "law2018_SD2017"
p2name <- "2018 law without standard deduction increase"
```


```{r 2018vsSD2017_summary_allstates}

tab_title <- "Total federal income tax liability in 2018: 2018 law, and 2018 law without standard deduction increase"
tab_subtitle <- "Amounts are in billions of 2018 dollars"

data <- pol_change(allsums_wide, policy1, policy2) %>%
  filter(agi_stub==99) %>%
  select(all_of(idvars), policy1, policy2, diff, pdiff) %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr) %>%
  select(stname, policy1, policy2, diff, pdiff)

tab <- pol_change_table(data, tab_title, tab_subtitle, p1name, p2name)
gtsave(tab, "2018vsSD2017_summary_table.png", path = here::here("results"))
tab

```


```{r 2018vsSD2017_one_state_byagi}

dollars <- "millions"

for(st in c("US", "NY")){  # nyus_sort
  print(st)
  tab <- get_state_tab(st, policy1, policy2, p1name, p2name, dollars)
  gtsave(tab, paste0("2018vsSD2017_byagi_table_", st, ".png"), path = here::here("results"))
  # print(tab)
}


```


```{r 2018vsSD2017_pdiff_byagi_wide}

tab_title <- "Total federal income tax liability in 2018: 2018 tax law, and 2018 law without standard deduction increase"
tab_subtitle <- "Percent change from 2018 law to 2018 law without standard deduction increase"

data <- pol_change(allsums_wide, policy1, policy2) %>%
  select(all_of(idvars), agi_stub, agi_label, stname, pdiff) %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr, agi_stub) %>%
  select(agi_label, stname, pdiff) %>%
  pivot_wider(names_from = stname, values_from = pdiff)

tab <- pdiff_wide_table(data, tab_title, tab_subtitle)
gtsave(tab, "2018vsSD2017_pdiff_byagi_wide_table.png", path = here::here("results"))
tab

```

## Impact of standard deduction combined with uncapped SALT, vs 2018

```{r}
policy1 <- "law2018"
p1name <- "2018 law"

policy2 <- "law2018_SALTSD2017"
p2name <- "2018 law with 2017 SALT and standard deduction"

tab_title <- "Total federal income tax liability in 2018: 2018 law with and without SALT and SD"

```


```{r 2018vsSALTSD2017_summary, rows.print=20}
# Don't forget to update policies and names!!!
tab_subtitle <- "Amounts are in billions of 2018 dollars"

data <- pol_change(allsums_wide, policy1, policy2) %>%
  filter(agi_stub==99) %>%
  select(all_of(idvars), policy1, policy2, diff, pdiff) %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr) %>%
  select(stname, policy1, policy2, diff, pdiff)

tab <- pol_change_table(data, tab_title, tab_subtitle, p1name, p2name)
gtsave(tab, "2018vsSALTSD2017_summary_table.png", path = here::here("results"))
tab

```


```{r 2018vsSALTSD2017_one_state_byagi}

dollars <- "millions"

for(st in c("US", "NY")){  # nyus_sort
  print(st)
  tab <- get_state_tab(st, policy1, policy2, p1name, p2name, dollars)
  gtsave(tab, paste0("2018vsSALTSD2017_byagi_table_", st, ".png"), path = here::here("results"))
  # print(tab)
}


```



```{r 2018vsSALTSD2017_pdiff_byagi_wide, rows.print=20}

tab_subtitle <- "Percent change from 2018 law to 2018 law without SALT and SD"

data <- pol_change(allsums_wide, policy1, policy2) %>%
  select(all_of(idvars), agi_stub, agi_label, stname, pdiff) %>%
  mutate(stabbr=factor(stabbr, levels=usny_sort)) %>%
  arrange(stabbr, agi_stub) %>%
  select(agi_label, stname, pdiff) %>%
  pivot_wider(names_from = stname, values_from = pdiff)


tab <- pdiff_wide_table(data, tab_title, tab_subtitle)

gtsave(tab, "2018vsSALTSD2017_pdiff_byagi_wide_table.png", path = here::here("results"))
tab

```

# Walk from 2017 to 2018, using diffs from 2018

```{r}
# what i want is a table with total liability in 2018 by income range, then change from policy 1, change from policy 2, ...,
# ...., synergy, 2017 law```{r ONETIME_long}

pid_stubs <- read_parquet(paste0(tcout, tc2017), 
                         col_select = all_of(c("pid", "c00100"))) %>%
  mutate(agi_stub=get_stub(c00100, agi_bounds))

keepvars <- c("pid", "iitax")

getsums <- function(fname, pname){
  df <- read_parquet(paste0(tcout, fname),
                         col_select = all_of(keepvars)) %>%
    mutate(policy=pname) %>%
    left_join(pid_stubs %>% select(pid, agi_stub), by="pid") %>%
    left_join(weights_long, by="pid")
  df2 <- df %>%
    group_by(agi_stub, stabbr, policy) %>%
    summarise(iitax=sum(weight * iitax), .groups="drop")
  return(df2)
}

dfall <- bind_rows(getsums(tc2017, "law2017"),
                   getsums(tc2018xSALT, "law2018_xSALT"),
                   getsums(tc2018_SD2017, "law2018_SD2017"),
                   getsums(tc2018_SALTSD2017, "law2018_SALTSD2017"),
                   getsums(tc2018, "law2018"))
                   
# dfall <- dfall %>%
#   left_join(agistubs %>% select(agi_stub, agi_label),
#             by="agi_stub")

dfsums <- dfall %>%
  group_by(policy, stabbr) %>%
  summarise(across(.cols=-agi_stub, ~sum(.x)), .groups="drop") %>%
  mutate(agi_stub=99)

dfallsums <- bind_rows(dfall, dfsums) %>%
  left_join(agistubs %>% select(-c(agi_ge, agi_lt)), by="agi_stub") %>%
  left_join(stcodes %>% select(stabbr, stname), by="stabbr") %>%
  mutate(stname=ifelse(stabbr=="other", "Other states", stname)) %>%
  arrange(policy, stabbr, agi_stub) %>%
  select(policy, stabbr, stname, agi_stub, agi_label, everything())

# define the order desired
order <- c("law2017", "law2018_xSALT", "law2018_SD2017", "law2018")
dfwide <- dfallsums %>%
  filter(policy != "law2018_SALTSD2017") %>%
  mutate(policy=factor(policy, levels = order)) %>%
  pivot_wider(names_from = policy, values_from = iitax)

data <- dfwide %>%
  filter(stabbr=="NY") %>%
  mutate(totchange=law2018 - law2017,
         saltchange=law2018 - law2018_xSALT,
         sdchange=law2018 - law2018_SD2017,
         other=totchange - (saltchange + sdchange)) %>%
  arrange(agi_stub) %>%
  select(agi_label, law2017, sdchange, saltchange, other, totchange, law2018)

data %>%
  kable(format.args = list(big.mark=","))


```

```{r}
xwalk_table <- function(data, tab_title, tab_subtitle){
  data %>%
    gt() %>%  tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      agi_label = "Adjusted gross income"
    ) %>%
    fmt_percent(
      columns=2:ncol(data),
      scale_values = FALSE,
      decimals = 1
    ) %>% 
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_source_note(source_note)
}

tab_title <- "xwalk table"
tab_subtitle <- "$ billions"
dollar_decimal <- 1
dollar_scale <- 1e-9
# law2017	sdchange	saltchange	other	law2018	totchange

tab <- data %>%
    gt() %>%  
      # cols_move(
      #   columns = vars(totchange),
      #   after = vars(law2018)
      #   ) %>%
    tab_header(
      title = tab_title,
      subtitle = tab_subtitle
    ) %>%
    cols_label(
      agi_label = "Adjusted gross income",
      law2017 = html("2017 law"),
      sdchange = html("Standard<br>deduction"),
      saltchange = html("SALT cap"),
      other = html("Other<br>changes"),
      law2018 = html("2018 law"),
      totchange = html("Total<br>change")
    ) %>%
    tab_spanner(
      label = "Tax change, measured in isolation relative to 2018 law, attributable to:",
      columns = vars(sdchange, saltchange, other)
    ) %>%
    tab_style(
      style = list(
        cell_borders(sides = "right", color = "#000000", style = "solid", weight = px(0.5))
      ),
      locations =list(
        cells_body(columns = vars(law2017, totchange)),
        cells_column_labels(columns = vars(law2017, totchange))
      )
    ) %>%
    tab_style(
      style = list(
        cell_borders(sides = "left", color = "#000000", style = "solid", weight = px(0.5))
      ),
      locations =list(
        cells_body(columns = vars(totchange)),
        cells_column_labels(columns = vars(totchange))
      )
    ) %>%    
    fmt_currency(
      columns = one_of(setdiff(names(data), "agi_label")),
      rows=1,
      decimals = dollar_decimal,
      scale_by = dollar_scale,
      suffixing = FALSE
    ) %>%
    fmt_number(
      columns = one_of(setdiff(names(data), "agi_label")),
      rows=2:nrow(data),
      decimals = dollar_decimal,
      scale_by = dollar_scale,
      suffixing = FALSE
    ) %>%
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_source_note(source_note)

gtsave(tab, "2017to2018_xwalk.png", path = here::here("results"))


```

